---
title: "A graph theoretic approach to constructing Room squares"
author: "Matthew Henderson"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    code_folding: hide
    fig_caption: yes
    theme: paper
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

```{r setup2, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

Graph factorisations
--------------------

A graph $G(V,E)$ consists of two sets. The first $V$, is called the
vertex-set, while the other $E$ consists of unordered pairs of $V$ and
is called the edge set. Usually graphs are represented with diagrams
where the members of $V$ are drawn as points and the members of $E$ as
lines connecting points. Adjacency for two vertices means being
connected by an edge. The **complete graph** $K_n$ is the graph on $n$
vertices in which all distinct vertices are adjacent.

```{r, echo = FALSE, fig.cap = "$K_4$ and $K_5$"}
library(igraph)
g <- make_full_graph(4)
h <- make_full_graph(5)

par(mfrow = c(1, 2))

plot(g, layout = layout_in_circle(g), vertex.color = "white", vertex.size = 20)
plot(h, layout = layout_in_circle(h), vertex.color = "white", vertex.size = 20)
```

A **one-factor** $f_i$ is a set of edges in which each vertex appears
exactly once.

#### Example 2.1.1

Two possible one-factors of $K_4$ are:
$$f_1 = \{12,34\}, f_2 = \{13,24\}$$
**Figure 10 Two one-factors of $K_4$**

A **one-factorisation** of the complete graph is a set of one-factors
in which all possible edges (i.e. all unordered pairs from the edge-set)
appear exactly once.

```{r, echo = FALSE, fig.cap = "$K_6$"}
g <- make_full_graph(6)

par(mfrow = c(1, 1))

plot(g, layout = layout_in_circle(g), vertex.color = "white", vertex.size = 20)
```

#### Example 2.1.2

Here $G = K_6$ the complete graph on 6 vertices with
$$V = \{1, 2, 3, 4, 5, 6\}$$
$$E = \{12, 13, 14, 15, 16, 23, 24, 25, 26, 34, 35, 36, 45, 46, 56\}$$
The one-factors are
$$f_1 = \{12, 35, 46\} \hspace{0.5cm} f_2 = \{14, 23, 56\} \hspace{0.5cm} f_3 = \{16, 25, 34\} 
\hspace{0.5cm} f_4 = \{13, 26, 45\} \hspace{0.5cm}  f_5 = \{15, 24, 36\}$$
because $f_1 \cup f_2 \cup f_3 \cup f_4 \cup f_5 = E$
$F = \{f_1, f_2, f_3, f_4, f_5\}$ is a one-factorisation of $G$ shown in
Figure 11.

```{r, fig.cap = "One-factorisation of $K_6$"}
E(g)[c(1, 11, 14)]$color <- "red"
E(g)[c(3, 6, 15)]$color <- "blue"
E(g)[c(5, 8, 10)]$color <- "green"
E(g)[c(2, 9, 13)]$color <- "yellow"
E(g)[c(4, 7, 12)]$color <- "black"

plot(g, layout = layout_in_circle(g), vertex.color = "white", vertex.size = 20)
```

Two one factors $f$ and $l$ are said to be ***orthogonal*** if
$f \cap l$ contains at most one edge. Two one-factorisations $F$ and $L$
are orthogonal if every one-factor in $F$ is orthogonal to every
one-factor in $L$.

Once again consider the square array in Figure 2. If the individual
elements within the array constituted the vertex set of a graph (call it
$R$) and the pairs within each box of the array were edges, we know that
each row is a one-factor and each column is a one-factor (because each
member of $R$ occurs precisely once in each row and once in each
column). Further more, because all edges from the edge-set of the
complete graph (i.e. all unordered pairs from $R$) appear once within
the array, we know that the rows together form a one-factorisation and
the columns form another, different, one-factorisation of $K_8$. Also,
because any row factor intersects any column factor in only one pair
(edge), all the row factors are orthogonal to all the column factors and
hence the two one-factorisations are orthogonal. We have demonstrated
the following theorem, given in [8] and proven in [19].

#### Theorem 2.1.1

The existence of a Room square of side $n$ is equivalent to the
existence of two orthogonal one-factorisations of the complete graph
$K_{n+1}$.

An example is given in figure 12 based on the Room square in Figure 2.

**Figure 12 Two orthogonal one-factorisations of $K_8$ based on
Kirkman’s square of 1850.**

Hill-Climbing Algorithm for Room Squares
----------------------------------------

The idea behind Hill-climbing algorithms is to suppose there exists a
**neighbourhood** of feasible solutions to some problem **instance**. With
each **feasible** solution there is an associated **cost** (or profit) and
finding an optimal solution becomes a matter of finding the solution with
minimum cost (or maximum profit).

A hill-climbing algorithm non-deterministically selects a solution from
the neighbourhood system such that the cost is less than that of some
initial solution until its procedure fails, hence finding the locally
optimal solution.

### An algorithm for One-Factorisations

Consider how to find a one-factorisation of the complete graph. Here the
problem instance is simply the even integer $n$ and vertex set $V$.\
Recall:

-   A one-factor of $K_n$ is a set of $n/2$ edges (hence $n$ is even)
    which partition $V$.

-   A one-factorisation of $K_n$ is a set of $n-1$ one-factors which
    partitions the edge set of $K_n$.

Suppose we choose to represent a one-factorisation by a set of
$\frac{n}{2}(n-1)=(n^2-n)/2$ pairs each of the form $(f_i,\{x,y\})$,
where $x \neq y, i=1...n-1$, and the following two conditions hold.

1.  Every $\{x,y\}$ occurs in a unique pair $(f_i,\{x,y\})$.

2.  For every one-factor $f_i$ and every vertex $x$, there is a unique
    pair of the form $(f_i,\{x,y\})$.

where $f_i$s are one-factors.

Then we consider a feasible solution to be a partial one-factorisation,
again represented by pairs having the same form but this time,

1.  Every $\{x,y\}$ occurs in at most one pair $f_i,\{x,y\})$.

2.  For every one-factor $f_i$ and every vertex $x$ there is at most one
    pair of the form $(f_i,\{x,y\})$.

Where the $f_i$s are **partial one-factors**.

Which enables a definition for the cost of a feasible solution $F$ to be
given by: $$c(F) = (n^2-n)/2-|F|$$ So that $F$ is a one-factorisation if
and only if $c(F)=0$, i.e. $|F|=(n^2-n)/2$

Now suppose that we can implement some procedure $X$, say, which either
reduces the cost or leaves it unaffected (i.e. it never increases the
cost) then the following "hill-climbing" algorithm, provided it
terminates, will find a one-factorisation.

  --------------------- --
  While $c(F) \neq 0$   
  Do $X$                
  --------------------- --

A procedure such as $X$ is called a heuristic. The following two
heuristics (due to Dinitz & Stinson [9]) when used together are suitable
for finding a one-factorisation.

Let $F$ be a partial one-factorisation of $K_n$:

Heuristic $H_1$ [8]

1.  Choose any vertex $x$ such that $x$ does not occur in every partial
    one-factor of $F$ (such a vertex is said to be a **live point**).

2.  Choose any partial one-factor $f_i$ such that $x$ does not occur in
    $f_i$.

3.  Choose any $y \neq x$ such that there is no partial one-factor $f_j$
    for which $(f_j,\{x,y\}) \in F$ (we say that $x$ and $y$ *do not
    occur together*).

4.  **if** $y$ does not occur in $f_i$, **then**

5.  $\hspace{1cm}$ Replace $F$ with $F \cup \{(f_i,\{x,y\})\}$.

6.  **Else** there is a pair in $F$ of the form
    $(f_i,\{z,y\}) \hspace{0.5cm} (z \neq x)$

7.  $\hspace{1cm}$ Replace $F$ with $F \cup \{(f_i,\{x,y\})\} \backslash
     \{(f_i,\{z,y\})\}$.

Heutistic $H_2$ [8]

1.  Choose any partial one-factor $f_i$ which does not occur in exactly
    $n/2$ pairs in $F$ (such a partial one-factor is said to be
    ***live***).

2.  Choose any $x$ and $y$ such that $x$ and $y$ do not occur together
    in $f_i$.

3.  **if** $x$ and $y$ do not occur together, **then**

4.  $\hspace{1cm}$ Replace $F$ with $F \cup \{(f_i,\{x,y\})\}$.

5.  **Else** there is a pair in $F$ of the form
    $(f_j,\{x,y\}) \hspace{0.5cm} (j \neq i)$

6.  $\hspace{1cm}$ Replace $F$ with $F \cup \{(f_i,\{x,y\})\} \backslash
     \{(f_j,\{x,y\})\}$.

#### Example

Suppose we are in the process of trying to find a one-factorisation for
$K_6$, and have generated a partial one-factorisation represented by the
set $F$.
$$F=\{(f_1,\{4,6\}),(f_1,\{3,5\}),(f_2,\{5,6\}),(f_3\{1,6\}), (f_3\{3,4\}),(f_4,\{2,3\}),(f_4,\{4,5\})\}$$
Now apply $H_1$:

1.  Choose $x=2$. Live, because it doesn’t appear in $f_1,f_2,f_3$ or
    $f_5$.

2.  Of these four partial one factors, choose $f_1$.

3.  2 only occurs together with 3 (in $f_4$), so pick $y=5$.

4.  5 already appears in $f_1$ so $\{z,y\}=\{3,5\}$. So replace $F$ by
    $F \cup \{(f_1,\{2,5\}) \backslash (f_1,\{3,5\})\}$

So we have extracted one edge from the one-factorisation and replaced it
with another edge, leaving the cost unchanged.
If in 3. we had picked 1 then according to the heuristic we should
replace $F$ with $F \cup (f_1,\{2,1\})$, increasing $|F|$ by one, and so
decreasing the cost by the same. Because the cost cannot increase $H_1$
is a suitable heuristic for use in a hill-climbing algorithm.

Now apply $H_2$ to the new one-factorisation $F_1=F \cup (f_1,\{2,1\})$

1.  We can pick any of $f_2, f_3, f_4, f_5$, because all are live.
    Choose $f_2$.

2.  Choose $x=2, y=3$, because neither appear in $f_2$.

3.  2 and 3 occur together in $f_4$. So replace $F_1$ with
    $F_1 \cup \{(f_2,\{2,3\}) \backslash (f_4,\{2,3\})\}$

Again the cost remains unchanged by this procedure, and if in 2. we had
chosen $x=1,y=4$ instead then we would have replaced $F_1$ with
$F_1 \cup \{(f_2,\{1,4\})\}$ decreasing the cost by one. As with $H_1$,
the cost cannot increase, which makes $H_2$ a suitable heuristic.
The hill-climbing algorithm for constructing one-factorisations which
was first given in [9] has a very simple form.

1.  **While** $c(F) \neq 0$, **do**

2.  choose $r=1$ or $r=2$ with equal probability

3.  perform $H_r$

### An Algorithm for Room Squares

To generate a Room square all that remains is to produce another
one-factorisation $G$, say, which is orthogonal to $F$. This will
inevitably require slight modifications to be made to $H_1$ and $H_2$.
Now if an array $R$ is constructed in which the rows are labelled with
the one-factors of $F(f_1,f_2,...,f_{n-1})$, and the columns are
labelled with the partial one-factors of $G(g_1,g_2,...,g_{n-1})$. Then
$R$ will be a Room square if the $(f_i,g_j)$ cell contains $\{x,y\}$, if
and only if $(f_i,\{x,y\}) \in F$ and $(g_j,\{x,y\}) \in G$ and is empty
otherwise.

Again these two heuristics are due to Dinitz & Stinson and originally
presented in [9]. Although a necessary correction has been made as will
become apparent.

$OH_1$

1.  Choose any live point $x$.

2.  Choose any partial one-factor $g_i$ such that $x$ does not occur in
    $g_i$.

3.  Choose any $y \neq x$ such that $x$ and $y$ do not occur together in
    $G$.

4.  Let $f_j$ be the one-factor of $F$ which contains the edge
    $\{x,y\}$.

5.  **if** $R(f_j,g_i)$ is not empty **then**

6.  $OH_1$ fails.

7.  **Else if** $y$ does not occur in $g_i$, **then**

8.  $\hspace{1cm}$ Replace $G$ by $G \cup (g_i,\{x,y\})$.

9.  $\hspace{1cm}$ Define $R(f_j,g_i)=\{x,y\}$.

10. **Else** there is a pair in $G$ of the form
    $(g_i,\{z,y\}) \hspace{0.5cm} z \neq x$.

11. $\hspace{1cm}$ Replace $G$ by
    $G \cup (g_i,\{x,y\}) \backslash (g_i,\{z,y\})$.

12. $\hspace{1cm}$ Define $R(f_k,g_i)$, to be empty$^i$, where
    $(f_k,\{z,y\}) \in F$.

$OH_2$

1.  Choose any live partial one-factor $g_i$.

2.  Choose any $x$ and $y \neq x$ such that $x$ and $y$ do not occur
    together in $g_i$.

3.  Let $f_j$ be the one-factor of $F$ which contains the edge
    $\{x,y\}$.

4.  **if** $R(f_j,g_i)$ is not empty **then**

5.  $OH_2$ fails.

6.  **Else if** $x$ and $y$ do not occur together, **then**

7.  $\hspace{1cm}$ Replace $G$ by $G \cup (g_i,\{x,y\})$.

8.  $\hspace{1cm}$ Define $R(f_j,g_i)=\{x,y\}$.

9.  **Else** there is a pair in $G$ of the form
    $(g_k,\{x,y\}) \hspace{0.5cm} (k \neq i$)

10. $\hspace{1cm}$ Replace $G$ by
    $G \cup (g_i,\{x,y\}) \backslash (g_k,\{x,y\})$

11. $\hspace{1cm}$ Define $R(f_j,g_i)=\{x,y\}$

12. $\hspace{1cm}$ Define $R(f_j,g_k)$ to be empty

#### *Example 2.2.1*

Suppose the factorisation $F$ from the earlier example has been
completed and is represented by the set:

$$ M={

  ----------------------------------------------------------------------------------------------
  $(f_1\{1,2\}),(f_1\{3,5\}),(f_1\{4,6\}),(f_2\{1,4\}),(f_2\{2,3\}),(f_2\{5,6\}),(f_3\{1,6\})$
  $(f_3\{2,5\}),(f_3\{3,4\}),(f_4\{1,3\}),(f_4\{2,6\}),(f_4\{4,5\}),(f_5\{1,5\}),(f_5\{3,6\})$
  ----------------------------------------------------------------------------------------------

}

Notice that this is precisely the one-factorisation of $K_6$ given
on page 9.

Now suppose we have established the following one-factors in $G$:
$$G=\{(g_1,\{1,4\}),(g_2,\{1,6\}),(g_3,\{3,6\}),(g_5\{5,6\}),(g_5\{1,2\})\}$$
At this state $R$ looks like 

$$R=

           $g_1$   $g_2$   $g_3$   $g_4$   $g_5$
  ------- ------- ------- ------- ------- -------
   $f_1$                                   $1,2$
   $f_2$   $1,4$                           $5,6$
   $f_3$           $1,6$                  
   $f_4$                                  
   $f_5$                   $3,6$          

$$

**Figure 13**

Now apply $OH_1$:

1.  Choose $x=5$, suitably live.

2.  Choose $g_3$, in which $5$ does not occur.

3.  $5$ does not occur together with $2$ in $G$, so we are free to
    choose $y=2$.

4.  In $F$, $\{2.5\} \in f_3$.

5.  $f_3,g_3$ is empty in $R$, also $y=2 \notin g_3$.

6.  Replace $G$ with $G \cup (g_3,\{5,2\})$.

7.  Define $R(f_3,g_3)=\{5,2\}$.

This decreases the cost by one, alternatively we might have chosen, at
stage 3. $y=3$, in that case.

4.  $\{3,5\} \in f_1$.

5.  $f_1,g_3$ is empty in $R$, also $y \in g_3$, occurring in the pair
    $(g_3,\{3,6\}), z=6$.

6.  Replace $G$ with $G \cup (g_3,\{3,5\}) \backslash (g_3,\{3,6\})$.

7.  Define $R(f_1,g_3)=\{3,5\}$.

8.  Define $R(f_5,g_3)$ to be empty.

Which leaves the cost unaffected. Suppose now that $R$ is the array
after this second version of the application of $OH_1$:

$$R=

           $g_1$   $g_2$   $g_3$   $g_4$   $g_5$
  ------- ------- ------- ------- ------- -------
   $f_1$                   $3,5$           $1,2$
   $f_2$   $1,4$                           $5,6$
   $f_3$           $1,6$                  
   $f_4$                                  
   $f_5$                                  

$$

**Figure 14**

Now if we apply $OH_2$:

1.  Choose $g_4$, a live partial one-factor.

2.  Choose $x=1, y=2$, neither of which occur in $g_4$.

3.  $(f_1,\{1,2\}) \in F$.

4.  $f_1,g_4$ is empty in $R$, also $x$ and $y$ do occur together,
    $(g_5,\{1,2\}) \in G$.

5.  Replace $G$ with $G \cup (g_4,\{1,2\}) \backslash (g_5,\{1,2\})$

6.  Define $R(f_1,g_4)=\{1,2\}$

7.  Define $R(f_1,g_5)$ to be empty.

This procedure leaves the cost unaffected and if instead we had chosen
at $2$. $x=3,y=4$, then would have been required to replace $G$ with
$G \cup (g_4,\{3,4\})$, and put $\{3,4\}$ in cell $(f_3,g_4)$ of $R$, an
action which reduces the cost by one. However, we know that two
orthogonal one-factorisations of $K_6$ are equivalent to a Room square
of side 5, which has been shown not to exist. Hence it would be futile
to continue with this method in this particular case. Nevertheless the
example shows how the heuristics work.

There is no guarantee of success with repeated use of these heuristics,
although Dinitz & Stinson are quick to point out that the algorithm
involving $H_1$ and $H_2$ has never[^3] failed to produce the desired
one-factorisation. If we hope to use the $OH_1$ and $OH_2$ in a similar
algorithm then the possibility of failure becomes a real possibility.
Two possibilities exist, either both heuristics fail or successive use
of them leads to an infinite loop. In order to avoid both we introduce a
***threshold*** function, which simply arrests the progress of the
algorithm after a certain number of iterations of the heuristics. Dinitz
& Stinson found after experimentation that the following function is
suitable. $$T(n)=100n$$ Then the hill-climbing algorithm for finding a
Room square is as follows [8]:

1.  Use the previous hill climbing algorithm to construct $F$, a
    one-factorisation of $K_n$.

2.  Number of iterations initialised to be $0$

3.  While (number of iterations $<T(n)$) and $c(G) \neq 0$, do

4.  $\hspace{1cm}$ Choose $r=1$ or $r=2$ at random with equal
    probability

5.  $\hspace{1cm}$ Perform $OH_r$

6.  $\hspace{1cm}$ Increment number of iterations

### The Room Square Generator

Dinitz and Stinson choose to implement the above algorithm in Pascal,
and ran in on an Amdahl 5850 workstation. It was very successful,
finding many Room squares with sides ranging from 11 to 101. For each
successful trial they had 9 or 10 failures (the program being stopped by
the threshold function) and timings ranged from 0.09 seconds for an
11x11 Room square, to 7.3 on average for the 25 different 101x101 Room
squares they found.

I chose to implement the hill-climbing algorithms in Visual Basic 6.0 on
a Pentium III-450/Win 98 Desktop. Needless to say, it was slightly less
successful – exhibiting a similar probability of success but
unfortunately becoming very slow for Room squares bigger than 21. It
found square of side 21 after an all-night search, but after 48 hours
looking for one of 23x23 I decided to call the search off.

Despite the failures at higher order, the Room square generator was very
successful in finding smaller squares. It found 7x7 Room squares in as
little as 4 seconds, and even 15x15 squares only took a few minutes.

Annotated code for the Room square generator can be found along with
some of the larger squares in Appendix I and below is a screen shot of
the application having successfully located a 9x9 Room square in a
little over one minute after 507 iterations of the heuristics $OH_1$ and
$OH_2$. The uppermost panel represents some of the one-factorisation
generated by the algorithm involving $H_1$ and $H_2$, while the second
panel shows part of the orthogonal one-factorisation generated by $OH_1$
and $OH_2$. The lower panel is a Room square of side 9.

**Figure 15 Screenshot of the Room Square Generator**
