---
title: "Hill-Climbing for One-Factorisations"
author: "Matthew Henderson"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    code_folding: show
    fig_caption: yes
    theme: paper
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

```{r setup2, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

http://kateto.net/netscix2016

https://jameshoward.us/writing/computational-methods-numerical-analysis-r/

http://www.bnlearn.com/
http://www.bnlearn.com/documentation/man/hc.html

## One-factors

We begin with a complete graph on six vertices. In **igraph** such a graph can
be generated with the `make_full_graph()` function.

```{r}
library(igraph)

g <- make_full_graph(6)
plot(g)
```

Next we give the vertices and edges names, to make them easier to refer to.

```{r}
V(g)$name <- LETTERS[1:6]
E(g)$name <- 1:15
```

```{r}
library(ggraph)

ggraph(g, layout = 'kk') + 
  geom_node_text(aes(label = name), size = 4) +
  geom_edge_link(aes(label = name), show.legend = FALSE, angle_calc = 'along', label_dodge = unit(3.0, 'mm'), label_push = unit(-6.0, 'mm'), start_cap = circle(4, 'mm'), end_cap = circle(4, 'mm'), label_colour = "blue") +
  theme_graph(foreground = 'steelblue', fg_text_colour = 'white')
```

A **one-factor** $f_i$ is a set of edges in which each vertex appears
exactly once.

For example,

```{r}
f1 <- c(1, 11, 14)
E(g)[f1]
```

```{r}
E(g)[f1]$factor <- "f1"

ggraph(g, layout = 'kk') + 
  geom_node_text(aes(label = name), size = 4) +
  geom_edge_link(aes(label = name, color = factor), show.legend = FALSE, angle_calc = 'along', label_dodge = unit(2.5, 'mm'), label_push = unit(-6.0, 'mm'), start_cap = circle(4, 'mm'), end_cap = circle(4, 'mm'), label_colour = "blue") +
  theme_graph(foreground = 'steelblue', fg_text_colour = 'white')
```

We need a function to test whether a set of edges constitutes a one-factor or
not. The `is_one_factor(f, g)` function will test whether or not $f$ is a one-factor
of $g$ or not by constructing a graph from the edges of $f$ and testing whether
this graph has the same vertex set as $g$ or not.

```{r}
is_one_factor <- function(f, g) {
  all(V(graph_from_edgelist(ends(g, f)))==V(g))
} 
```

Using this function we can test, for example, whether $f1$ is a one-factor of
$g$ or not.

```{r}
is_one_factor(f1, g)
```

## One-factorisations

A one-factorisation is a decomposition of the edge-set of a graph into
one-factors.

We can create a one-factorisation that includes $f1$ as one of its one-factors
by creating another four one-factors ($f2$, $f3$, $f4$, and $f5$ below).

```{r}
f2 <- c(3, 6, 15)
f3 <- c(5, 8, 10)
f4 <- c(2, 9, 13)
f5 <- c(4, 7, 12)

E(g)[f2]$factor <- "f2"
E(g)[f3]$factor <- "f3"
E(g)[f4]$factor <- "f4"
E(g)[f5]$factor <- "f5"
```

We can test that all of these are, indeed, one-factors.

```{r}
all(sapply(list(f2, f3, f4, f5), function(x) is_one_factor(x, g)))
```

Colouring the edges according to which factor they belong to makes it easy to
confirm that the one-factors together constitute a one-factorisation.

```{r}
gg1 <- ggraph(g, layout = 'kk') + 
  geom_node_text(aes(label = name), size = 4) +
  geom_edge_link(aes(label = name, color = factor), show.legend = FALSE, angle_calc = 'along', label_dodge = unit(2.5, 'mm'), label_push = unit(-6.0, 'mm'), start_cap = circle(4, 'mm'), end_cap = circle(4, 'mm'), label_colour = "blue") +
  theme_graph(foreground = 'steelblue', fg_text_colour = 'white')

print(gg1)
```

We can also plot the factors as separate graphs.

```{r, fig.width = 12, fig.height = 8}
gg1 + facet_edges(~factor)
```

```{r}
F1 <- list(f1, f2, f3, f4, f5)
F1
```

```{r}
is_one_factorisation <- function(of, g) {
  length(F1) == vcount(g) - 1 && all(sapply(of, is_one_factor, g)) && length(difference(E(g), E(g)[unlist(of)])) == 0
}
```

```{r}
is_one_factorisation(F1, g)
```

## Heuristics

```{r}
f1 <- c(1, 11)
f2 <- c(3, 6, 15)
f3 <- c(10)
f4 <- c(2, 9, 13)

pof <- list(f1, f2, f3, f4)
```


### Heuristic $H_{1}$

1.  Choose any vertex $x$ such that $x$ does not occur in every partial
    one-factor of $F$ (such a vertex is said to be a **live point**).

2.  Choose any partial one-factor $f_i$ such that $x$ does not occur in
    $f_i$.

3.  Choose any $y \neq x$ such that there is no partial one-factor $f_j$
    for which $(f_j,\{x,y\}) \in F$ (we say that $x$ and $y$ *do not
    occur together*).

4.  **if** $y$ does not occur in $f_i$, **then**

5.  $\hspace{1cm}$ Replace $F$ with $F \cup \{(f_i,\{x,y\})\}$.

6.  **Else** there is a pair in $F$ of the form
    $(f_i,\{z,y\}) \hspace{0.5cm} (z \neq x)$

7.  $\hspace{1cm}$ Replace $F$ with $F \cup \{(f_i,\{x,y\})\} \backslash
     \{(f_i,\{z,y\})\}$.

### Heutistic $H_{2}$

1.  Choose any partial one-factor $f_i$ which does not occur in exactly
    $n/2$ pairs in $F$ (such a partial one-factor is said to be
    ***live***).

2.  Choose any $x$ and $y$ such that $x$ and $y$ do not occur together
    in $f_i$.

3.  **if** $x$ and $y$ do not occur together, **then**

4.  $\hspace{1cm}$ Replace $F$ with $F \cup \{(f_i,\{x,y\})\}$.

5.  **Else** there is a pair in $F$ of the form
    $(f_j,\{x,y\}) \hspace{0.5cm} (j \neq i)$

6.  $\hspace{1cm}$ Replace $F$ with $F \cup \{(f_i,\{x,y\})\} \backslash
     \{(f_j,\{x,y\})\}$.

#### Example

Suppose we are in the process of trying to find a one-factorisation for
$K_6$, and have generated a partial one-factorisation represented by the
set $F$.
$$F=\{(f_1,\{4,6\}),(f_1,\{3,5\}),(f_2,\{5,6\}),(f_3\{1,6\}), (f_3\{3,4\}),(f_4,\{2,3\}),(f_4,\{4,5\})\}$$
Now apply $H_1$:

1.  Choose $x=2$. Live, because it doesn’t appear in $f_1,f_2,f_3$ or
    $f_5$.

2.  Of these four partial one factors, choose $f_1$.

3.  2 only occurs together with 3 (in $f_4$), so pick $y=5$.

4.  5 already appears in $f_1$ so $\{z,y\}=\{3,5\}$. So replace $F$ by
    $F \cup \{(f_1,\{2,5\}) \backslash (f_1,\{3,5\})\}$

So we have extracted one edge from the one-factorisation and replaced it
with another edge, leaving the cost unchanged.
If in 3. we had picked 1 then according to the heuristic we should
replace $F$ with $F \cup (f_1,\{2,1\})$, increasing $|F|$ by one, and so
decreasing the cost by the same. Because the cost cannot increase $H_1$
is a suitable heuristic for use in a hill-climbing algorithm.

Now apply $H_2$ to the new one-factorisation $F_1=F \cup (f_1,\{2,1\})$

1.  We can pick any of $f_2, f_3, f_4, f_5$, because all are live.
    Choose $f_2$.

2.  Choose $x=2, y=3$, because neither appear in $f_2$.

3.  2 and 3 occur together in $f_4$. So replace $F_1$ with
    $F_1 \cup \{(f_2,\{2,3\}) \backslash (f_4,\{2,3\})\}$

Again the cost remains unchanged by this procedure, and if in 2. we had
chosen $x=1,y=4$ instead then we would have replaced $F_1$ with
$F_1 \cup \{(f_2,\{1,4\})\}$ decreasing the cost by one. As with $H_1$,
the cost cannot increase, which makes $H_2$ a suitable heuristic.
The hill-climbing algorithm for constructing one-factorisations which
was first given in [9] has a very simple form.

1.  **While** $c(F) \neq 0$, **do**

2.  choose $r=1$ or $r=2$ with equal probability

3.  perform $H_r$


## Hill-climbing

There is a function `hill.climbing.search` in the FSelector package.

```{r}
library(rpart)
data(iris)
evaluator <- function(subset) {
#k-fold cross validation
k <- 5
splits <- runif(nrow(iris))
results = sapply(1:k, function(i) {
test.idx <- (splits >= (i - 1) / k) & (splits < i / k)
train.idx <- !test.idx
test <- iris[test.idx, , drop=FALSE]
train <- iris[train.idx, , drop=FALSE]
tree <- rpart(as.simple.formula(subset, "Species"), train)
error.rate = sum(test$Species != predict(tree, test, type="c")) / nrow(test)
return(1 - error.rate)
})
print(subset)
print(mean(results))
return(mean(results))
}

library(FSelector)
subset <- hill.climbing.search(names(iris)[-5], evaluator)
f <- as.simple.formula(subset, "Species")
print(f)
```













