--- 
title: "Room Squares"
author: "Matthew Henderson"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: mhenderson/room
description: "This is a minimal example of using the bookdown package to write a book. The output format for this example is bookdown::gitbook."
---

# Kirkman’s Schoolgirl Problem

In 1850 Thomas Penyngton Kirkman, an English mathematician from Bolton,
published the following problem in the *Lady’s and Gentleman’s Diary.*

> *Fifteen young ladies of a school walk out three abreast for seven
> days in succession: it is required to arrange them daily so that no
> two shall walk abreast more than once*

In solving this problem Kirkman discovered the following square array,
which he observed was a very "curious arrangement".

```{r, echo = FALSE, results='asis'}
t <- matrix(
  c(
    "", "", "", "hi", "kl", "mn", "op",
    "", "il", "mo", "", "np", "hk", " ",
    "", "no", "hl", "mp", "", "", "ik",
    "lp", "", "in", "ko", "hm", "", "",
    "im", "", "kp", "", "", "lo", "hn"
  ), ncol = 7, byrow = TRUE
)

pander::pandoc.table(t, style = "grid", caption = "Kirkman’s curious arrangement")
```

The curiosity of this square is that each of the letters h, i, k, l, m, n, o, p
occurs precisely once in every column and row, while in the entire
square each of the letters makes a pair with every other letter exactly
once. Kirkman was able to employ this square to solve his Schoolgirl
Problem. To each pair in the first column he added the element 1, to
each pair in the second column 2 and so on. In addition he introduced
the missing triple of numbers to each row. (e.g. row one has no elements
in any of the first three columns so the numbers 1,2 and 3 would not
appear hence he would add the triple 123). The seven rows of unique
triples then corresponded to seven days in which the elements,
corresponding to schoolgirls, were paired together exactly once
throughout the arrangement. Thereby solving the problem.

```{r, echo = FALSE, results = 'asis'}
t <- matrix(
  c(
    "Day 1", "123", "hi4", "kl5", "mn6", "op7",
    "Day 2", "147", "il2", "mo3", "np5", "hk6",
    "Day 3", "156", "no2", "hl3", "mp4", "ik7",
    "Day 4", "267", "lo2", "in3", "ko4", "hm5",
    "Day 5", "245", "io2", "kp3", "lo6", "hn7",
    "Day 6", "357", "ho2", "km2", "ln4", "ip6",
    "Day 7", "346", "ko2", "hp2", "io5", "lm7"
  ),
  ncol = 6,
  byrow = TRUE
)

pander::pandoc.table(t, caption = "Kirkman's Schoolgirl's Solution")
```

Kirkman was a notable mathematician who is often regarded as the
originator of the object in Figure 2, which has subsequently become
known as a Room square (after T.G. Room).

Tournaments
-----------

Suppose the English Football Association proposed hosting a new type of
international tournament to be staged as a one-off event in England.
This tournament would involve eight national sides competing in a league
that would be staged in various stadia around the country over two
weeks. The structure of the tournament would be such that every team
played every other team once, with the winner being the team which
accumulated most points in the manner of a normal football league (3
points for a win, 1 for a draw).

To know which matches need playing is simple. Suppose the eight invited
teams are:

```{r, echo = FALSE, results='asis'}
t <- matrix(
  c(
    'Argentina', 'England',
    'Brazil', 'France',
    'Columbia', 'Germany',
    'Denmark', 'Holland'
  ), ncol = 2,
  byrow = TRUE
)

pander::pandoc.table(t, caption = "Teams")
```

If we write matches as alphabetic pairs in the obvious way, (e.g. ab
denoting Argentina versus Brazil). The complete list of matches (the
match set, M) is simply all unordered pairs from team set, T:
$$T = \{a, b, c, d, e, f, g, h\}.$$ i.e. 

$$ 
M = \{
  ab,ac,ad,ae,af,ag,ah,bc,bd,be,bf,bg,bh,cd,
  ce,cf,cg,ch,de,df,dg,dh,ef,eg,eh,fg,fh,gh
\}
$$

It remains to be decided where and when the matches will be played.

The English F.A., for whatever reason (the financial cost of hosting
eight teams, for example), has imposed a time limit of two weeks on the
tournament. Realistically the treams can only manage to play on
alternate days so it is decided to have, in effect, seven different
"rounds" with each team competing once in each round. (Seven being
the smallest number of rounds because each team has to play seven
others).

For reasons of fairness the F.A. also demands the condition that each
team will play once at each stadium.
Can such a tournament exist?
Suppose the stadia used are the following:

```{r, echo = FALSE, results='asis'}
t <- matrix(
  c(
    'Wembley', 'Highbury',
    'Villa Park', 'Stadium of Light',
    'Stamford Bridge', 'Old Trafford',
    'St. James Park'
  ), ncol = 1,
  byrow = TRUE
)

pander::pandoc.table(t, caption = "Stadium")
```

Then figure 4 provides a match schedule which is suitable for such a
tournament.

```{r, echo = FALSE, results='asis'}
t <- data.frame(
  "1" = c("  ", "  ", "  ", "dh", "be", "ag", "cf"),
  "2" = c("  ", "bd", "fg", "  ", "  ", "ce", "ah"),
  "3" = c("  ", "eg", "ad", "bf", "ch", "  ", "  "),
  "4" = c("ab", "  ", "eh", "cg", "  ", "df", "  "),
  "5" = c("cd", "fh", "  ", "ae", "  ", "  ", "bg"),
  "6" = c("ef", "ah", "  ", "  ", "bg", "bh", "  "),
  "7" = c("gh", "  ", "bc", "  ", "af", "  ", "de")
)

colnames(t) <- 1:7
# Hack to give some row names (doesn't work with 1:7)
rownames(t) <- paste(1:7, " ")
pander::pandoc.table(t, caption = "Fixture List for an International Soccer League")
```

Looking along the rows, each team plays once in each round. Looking down
columns, each stadia hosts each team exactly once. And throughout the
tournament as a whole each pair from the original match list appears
exactly once, hence every team opposes every other team once. Figure 4
is another Room square of side 7. Alternatively, because the pairs are
made from a set containing 8 elements, we say that this is a Room square
of order 8.

T.G. Room, (1902-86)
--------------------

In 1955, Thomas Gerald Room, then Professor of Mathematics at the
University of Sydney, published a brief note in the Mathematical Gazette
entitled *A new type of magic square*.[20] In it he presented another
example of a square array with the same properties as Kirkman’s. This
square, Room explained, had been discovered as "a by-product of
another investigation". It was preceded in the note by a particularly
efficient statement of the properties of these squares, which have
subsequently been known by his name.

> "The problem is to arrange the $n(2n-1)$ symbols $rs$ (which is the
> same as $sr$) formed from all pairs of $2n$ digits such that in each
> row and each column there appear $n$ symbols (and $n-1$ blanks) which
> among them contain all $2n$ digits."

Room’s note went on to explain that while the trivial $n = 1$ Room square
exists[^1] , the non-existence of those with $n = 2$ (side 2) and $n = 3$
(side 5) is easily proven. Room considered the $n = 2$ proof so
straightforward that it was omitted from this note, while for the $n = 3$
case he made reference to a graph-theoretic proof.

Consider the $n = 2$ case, we are required to place all the pairs of 4
digits in a 3x3 array. If we choose to use the set of non-negative
integers $\{0, 1, 2, 3\}$, then we need to find somewhere to put each of
the pairs $\{01, 02, 03, 12, 13, 23\}$. That we can swap the rows and columns
of a Room square without damaging that square’s Room-ness is
self-evident. Therefore there is no less of generality in assuming that
a $3 \times 3$ Room square has the pair $\{0, 1\}$ in cell $(1, 1)$.

```{r, echo = FALSE, results = 'asis'}
t <- matrix(
  c(
    "01"
  ),
  ncol = 1,
  byrow = TRUE
)

pander::pandoc.table(t, caption = "")
```

If we hope to make this array a Room square we must place the pair
$\{2, 3\}$ in the first row, while to complete the first column we must
also place the same pair in either position $(1, 2)$ or $(1, 3)$, but each
pair is only allowed to appear once. So there is no Room square of side
3, order 4.

For the $n = 3$, a Room square of side 5, case we consider the following
array:

```{r, echo = FALSE, results = 'asis'}
t <- matrix(
  c(
    "01", "23", "45", "-", "-",
    "24", "-", "-", "-", "-",
    "35", "-", "-", "-", "-"
  ),
  ncol = 5,
  byrow = TRUE
)

pander::pandoc.table(t, caption = "")
```

There is no loss in generality in using this array because we can
reorder rows and columns to obtain the first row in this form and then
the first column must contain either the pairs $\{01, 24, 35\}$ or
$\{01, 25, 34\}$ and the latter can be converted into the former by the
permutation $(45)$[^2] which leaves the first row unchanged. We now show
that completion of this square is impossible. The pairs
$\{2,5\},\{3,4\}$ must appear somewhere in the array other than the
first three rows or columns. Also they must appear in separate
rows/columns to prevent a forced recurrence of $\{0, 1\}$. Suppose we put
$\{2, 5\}$ in $(4, 4)$ and $\{3, 4\}$ in $(5, 5)$, then we know that cells
$(4, 5)$ and $(5, 4)$ are empty, as the only pair which could legally go
in either would be $\{0, 1\}$.

Hence we know that cells $(4, 2), (4, 3), (5, 2), (5, 3)$ each contain pairs.
Take cell $(5, 2)$, it could only contain $\{0, 5\}$ or $\{1, 5\}$, and the
latter becomes the former under $(01)$, so assume it contains $\{0, 5\}$.
We are now forced to fill in the other cells to give the array in Figure
8.

```{r, echo = FALSE, results = 'asis'}
t <- matrix(
  c(
    "01", "23", "45", "-", "-",
    "24", "-", "-", "-", "-",
    "35", "-", "-", "-", "-",
    "-", "14", "03", "25", "-",
    "-", "05", "12", "-", "34"
  ),
  ncol = 5,
  byrow = TRUE
)

pander::pandoc.table(t, caption = "")
```

We still need to place the pairs $\{0, 2\}$ and $\{0, 4\}$, which cannot
be done because neither can appear in the second row and they cannot
both appear in the third row. Hence there is no Room square of side 5,
order 6.

The real significance of Room’s note was that mathematicians soon took
on the task of determining the spectra of Room squares (those values of
$n$ for which Room squares exist). Research which cumulated 19 years
later in the complete statement of the existence of Room squares, made
by W.D. Wallis, that:

> *"Room squares exist for all odd positive integer sides except 3
> and 5"* [24]

Proving this statement, which was suspected to be true from an early
stage, turned out to be protracted and difficult.

The most significant breakthrough came in 1968 when Stanton and Mullin
introduced the starter-adder method for constructing Room squares. This
method reduces the problem of constructing Room squares to the problem
of finding a certain type of initial row from which a Room square can be
developed straightforwardly.

In this work much emphasis will be placed upon the proof of the
existence of Room squares.

The Galois Field
----------------

Throughout this work much use will be made of a particular *finite
field*, known as the Galois field, denoted by $GF(p^n)$. Whenever $p^n$
is a prime (i.e. $n=1$) the Galois field is precisely the integers under
modulo p arithmetic, denoted $Z_p$. The Galois field has a number of
important properties which are used in many of the proofs that follow,
we introduce some of these now.

-   Every Galois field (every finite field in fact) has a *primitive
    element*. An element, $x$ say, is primitive in $GF(q)$ if
    $x^0,x^1,x^2,...,x^{q-1}$ are all the non-zero members of $GF(q)$.

### Example

$x=2$ is a primitive element in $GF(11)$ because,
$$x^0 = 1 \hspace{0.5cm} x^1 = 2 \hspace{0.5cm} x^2 = 4 \hspace{0.5cm} x^3 = 8 \hspace{0.5cm} x^4 = 5$$
$$x^5 = 10 \hspace{0.5cm} x^6 = 9 \hspace{0.5cm} x^7 = 7 \hspace{0.5cm} x^8 = 3 \hspace{0.5cm} x^9 = 6$$

-   It can be shown [3] that $x^{q-1}=1$ is always true for any $GF(q)$
    where $q$ is odd, and $x^i \neq 1$ for any $1 \leq i \leq q-1$

-   $x^{q-1}=1$ implies that
    $(x^{\frac{1}{2}(q-1)}-1)(x^{\frac{1}{2}(q-1)}+1)=0$, therefore
    either $x^{\frac{1}{2}(q-1)}=1$ or $x^{\frac{1}{2}(q-1)}=-1$.
    Clearly because of the previous remark, only the latter can be true.

-   If $b$ is a non-zero residue modulo $p$, then $b$ is a quadratic
    residue (or square) if $x^2 \equiv b(\textrm{mod } p)$ has
    solutions, otherwise $b$ is a quadratic non-residue (or non-square).
    So the non-zero squares are precisely the even powers of the
    primitive element, while the non-zero non-squares are the odd
    powers.

-   There are precisely $\frac{1}{2}(p-1)$ squares mod $p$, and
    $\frac{1}{2}(p-1)$ non-squares.

-   $-1$ is a square if $q \equiv 1(\textrm{mod } 4)$, but not a square
    for $q \equiv 3(\textrm{mod } 4)$

    -   $q \equiv 1(\textrm{mod } 4)$, then if $x^i$ is a square so is
        $-x^i$.

    -   $q \equiv 3(\textrm{mod } 4)$, then $x^i$ is a square $-x^i$ is
        a non-square.

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
