---
title: "Room Squares"
author: "Matthew Henderson"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  github_document:
    toc: yes
---
    
# Introduction

```{r test-main, child='17_03_12_introduction.Rmd'}
```

# A graph-theoretic approach to constructing Room squares

```{r test-main, child='17_03_13_graph_theoretic.Rmd'}
```

# Proving the existence of Room squares

```{r test-main, child='17_03_13_existence_proof.Rmd'}
```

# Balanced Room squares

```{r test-main, child='17_03_15_balanced_room_squares.Rmd'}
```

# Closing Remarks

The results we have established in the previous chapter regarding the
existence of balanced Room squares represent by no means the complete
story. Du and Hwang [10] have established the existence of $SSBS$ for
all prime powers $$q = 2^{\alpha}t+1, \alpha \geq 2, t \geq 3, t$$ odd.
Further, Anderson has shown that consequently the construction due
originally to Hwang, Kang and Yu but corrected in [2], allows us to
state the existence of the corresponding $BRS(2q+2)$ in one particular
case.

By far the most significant remaining result which has not been included
in the previous chapter is due to B.A. Anderson who proved that
$BRS(2^n)$ exist for all odd $n \geq 3$. His construction was based upon
the theory of finite geometry, an area which has also contributed
constructions for Room squares (the non-balanced kind). Other similar
geometrical constructions have been used to establish the existence of
$BRS(2^n)$, for $4 \leq n \leq 18$, $n$ even. The two smallest values of
$n \equiv 0($mod 4) for which the existence of a $BRS(n)$ remains in
doubt are 36 and 92. The first of these, along with many others, would
be established by the doubling construction if a $SSBS$ could be found
in $Z_{17}$. This remains one of the most significant open problems for
$BRS$, namely to establish the existence of $SSBS$ in $Z_n$ when $n$ is
a Fermat prime.

The link between graph theory and Room squares that was touched upon in
the second chapter has opened many avenues of research. Possibly the
most interesting of which is the existence of perfect Room squares. A
one-factorisation of $K_n$ is said to be perfect if the union of two of
its one-factors is a hamiltonian cycle of $K_n$. A perfect Room square
is one of side $n$ in which both row and column factorisations of
$K_{n+1}$ are perfect. Very little seems to be known about perfect
one-factorisations. Individual examples of perfect Room squares of side
11 have been constructed but no infinite classes have yet been found.

[^1]: The Room squre of side1 is just the single array element
    containing the pair $\{0,1\}$.

[^2]: This is cycle notation, and stands for the permutation
    $4 \rightarrow 5, 5 \rightarrow 4$.

[^3]: In over ten-million attempts, they claim

[^4]: Both Howell and Whitfield had previously found starters and
    adders, but the precise method used here due to Stanton & Mullin

[^5]: The starter in then first Lemma has pairs whose elements always
    sum to $2n-1$, while the second Lemma, because $a^{n-1}=-1$, has
    pairs which can be written in the form $(a^x,-a^x)$.

[^6]: In 1880 F.Landry showed
    $F_6=2^{64}+1=274177 \times 67280421310721$. In 1975 Brillhart and
    Morrison showed
    $F_7=2^{128}+1=59649589127497217 \times 5704689200685129054721$. In
    1981, Brent and Pollard found that\
    $2^{256}+1=1238926361552897 \times
    93461639715357977769163558199606896584051237541638188580280321$

[^7]: Wallis presented a Room square of side 257 a conference in 1973,
    completing his proof (different from the one presented here) of the
    existence of Room squares.

[^8]: This is a theorem, not proven here

[^9]: We could have used the same permutation in each case, but the
    general theorem (as will be introduced) insists they are different,
    as higher order multiplications require different permutations.
    Quintuplication, for example, needs at least two.

[^10]: This definition is extracted in its entirety from [8]

[^11]: Where the blocks are derived in the same way as for a BRS

[^12]: The $t=1$ case was proven by Stanton and Sprott (1964) and Parker
    (1963) prior to Schellenberg.

[^13]: because $q \equiv 3$(mod 4)

[^14]: $A(X) \cap A(Y) \neq \varnothing$ implies, as $(1+x) \neq 0$,
    that $-x^{2i} = -x^{2i-1}$ for some $i$, i.e. $1=x^{-1}$,
    $\therefore A(X) \cap A(Y) = \varnothing$

